Semantic Linearity in a Lazy Optimising Compiler

### 2024-02-11 icfp.tex
minor changes

\renewcommand{\DeltaLinearRelationLemma}{
\begin{lemma}[$\Delta$-bound to Linear]
% Any $\Delta$-variable with a fully-irrelevant usage environment can be made a linear variable as long as its irrelevant environment is dropped from the linear environment (and the unrestricted env. is updated accordingly).\\
If $\G,\x[\irr{\D}]; \irr{\D},\D' \vdash e : \vp$
then $\G[x/\irr{\D}]; \D',\xl \vdash e :\vp$.
\end{lemma}
}

\renewcommand{\LinearDeltaRelationLemma}{
  \begin{lemma}[Linear to $\Delta$-bound]
If $\G; \D',\xl \vdash e :\vp$
then $\G[\D/x],\xD; \D,\D' \vdash e : \vp$.
\end{lemma}
}

\renewcommand{\DeltaUnrestrictedRelationLemma}{
\begin{lemma}[Unrestricted and $\Delta$-bound]
%An unrestricted variable is equivalent to a $\D$-var with an empty usage environment.\\
$\G,\xo; \D \vdash e : \vp$ iff $\G,\x[\cdot]; \D \vdash e : \vp$
\end{lemma}
}

>>>>

\renewcommand{\DeltaLinearRelationLemma}{
\begin{lemma}[$\Delta$-bound to Linear]\label{lem:deltaone}
% Any $\Delta$-variable with a fully-irrelevant usage environment can be made a linear variable as long as its irrelevant environment is dropped from the linear environment (and the unrestricted env. is updated accordingly).\\
If $\G,\x[\irr{\D}]; \irr{\D},\D' \vdash e : \vp$
then $\G[x/\irr{\D}]; \D',\xl \vdash e :\vp$.
\end{lemma}
}

\renewcommand{\LinearDeltaRelationLemma}{
  \begin{lemma}[Linear to $\Delta$-bound]\label{lem:onedelta}
If $\G; \D',\xl \vdash e :\vp$
then $\G[\D/x],\xD; \D,\D' \vdash e : \vp$.
\end{lemma}
}

\renewcommand{\DeltaUnrestrictedRelationLemma}{
\begin{lemma}[Unrestricted and $\Delta$-bound]\label{lem:undelta}
%An unrestricted variable is equivalent to a $\D$-var with an empty usage environment.\\
$\G,\xo; \D \vdash e : \vp$ iff $\G,\x[\cdot]; \D \vdash e : \vp$
\end{lemma}
}

### 2024-02-11 language-v4/proofs/WHNFConvSoundness.tex
Restate theorems and lemmas
minor changes

\WHNFConvSoundness

@@

\item[Case:] $Alt\_$
\begin{tabbing}
    (1) $\G,\z[\irr{\D}];\irr{\D}, \D' \vdash_{alt} \_ \to e :^z_{\irr{\D}} \s \Rightarrow \vp$\\
    (2) $\G, \z[\irr{\D}];\irr{\D},\D' \vdash e : \vp$\\
    (3) $\irr{\D}$ is used through $z$\`since $\irr{\D}$ can't otherwise be used\\\`and is introduced in this alternative uniquely \\\`(since we have multi-tier proof irrelevance, i.e. $\irr{\irr{\D}}\neq\irr{\D}$)\\
    (4) $\G; \z[1], \D' \vdash e : \vp$\` by $\D \Rightarrow 1$ lemma (2,3)\\
    (5) $\G, \z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash e : \vp$\` by $1 \Rightarrow \D$ lemma (4)\\
    (6) $\G, \z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash_{alt} \_ \to e :^z_{\ov{\D_i}} \s \Rightarrow \vp$\` by $Alt\_$\\
\end{tabbing}

@@

\item[Case:] $AltN_{\textrm{Not WHNF}}$
\begin{tabbing}
    (1) $\G, \z[\irr{\D}]; \irr{\D}, \D' \vdash_{alt} K~\ov{\xo},\ov{y_i{:}_{1}\s_i}^n\to e :^z_{\irr{\D}} \s \Rightarrow \vp$\\
    (2) $\G,\z[\irr{\D}],\ov{\xo},\ov{y_j{:}_{\D_j}\s_j}^n; \irr{\D}, \D' \vdash e : \vp$\`by inv. on $AltN_{\textrm{Not WHNF}}$\\
    (3) $\ov{\D_j} = \ov{\lctag{\irr{\D}}{K_j}}$\`by inv. on $AltN_{\textrm{Not WHNF}}$\\
    Subcase $\irr{\D}$ is consumed through $z$\\
    (4) $\G,\ov{\xo},\ov{y_j{:}_{\D_j}\s_j}^n; \z[1], \D' \vdash e : \vp$\`by $\D \Rightarrow 1$ lemma (2,subcase)\\
    (5) $\G,\ov{\xo},\ov{y_j{:}_{\D_i}\s_j}^n; \z[1], \D' \vdash e : \vp$\`by $\ov{y}$ does not consume resources,\\\` and Weaken? (HELP move $\D_j$ to $\D_i$)\\
    (6) $\G,\ov{\xo},\ov{y_j{:}_{\D_i}\s_j}^n, \z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash e : \vp$\`by $1 \Rightarrow \D$ lemma (5)\\
    (7) $\G, \z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash_{alt} K~\ov{\xo},\ov{y_i{:}_{1}\s_i}^n \to e :^z_{\ov{\D_i}} \s \Mapsto \vp$\`by $AltN_{\textrm{WHNF}}$ (6)\\
    Subcase $\irr{\D}$ is (fully) consumed by $\ov{y}$ (after splitting)\\
    (4) $\G,\z[\irr{\D}],\ov{\xo};\ov{y_j{:}_{1}\s_j}^n, \D' \vdash e : \vp$\`by $\D \Rightarrow 1$ lemma (2, subcase)\\
    (5) $\G,\z[\ov{\D_i}],\ov{\xo};\ov{y_j{:}_{1}\s_j}^n, \D' \vdash e : \vp$\`by Weaken and $z$ does not consume resources (HELP SAME KIND)\\
    (6) $\G,\z[\ov{\D_i}],\ov{\xo},\ov{y_j{:}_{\D_i}\s_j}^n; \ov{\D_i}, \D' \vdash e : \vp$\`by $1 \Rightarrow \D$ lemma (5)\\
    (7) $\G,\z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash K~\ov{\xo},\ov{y_i{:}_{1}\s_i}^n\to e :^z_{\ov{\D_i}} \s \Mapsto \vp$\`by $AltN_{\textrm{WHNF}}$\\
\end{tabbing}

>>>>

\irrelevancelemma*

@@

\item[Case:] $Alt\_$
\begin{tabbing}
    (1) $\G,\z[\irr{\D}];\irr{\D}, \D' \vdash_{alt} \_ \to e :^z_{\irr{\D}} \s \Rightarrow \vp$\\
    (2) $\G, \z[\irr{\D}];\irr{\D},\D' \vdash e : \vp$\\
    (3) $\irr{\D}$ is used through $z$\`since $\irr{\D}$ can't otherwise be used\\\`and is introduced in this alternative uniquely \\\`(since we have multi-tier proof irrelevance, i.e. $\irr{\irr{\D}}\neq\irr{\D}$)\\
    (4) $\G[z/\irr{\D}]; \z[1], \D' \vdash e : \vp$\` by Lemma~\ref{lem:deltaone} (2,3)\\
    (5) $\G[\ov{\D_i}/\irr{\D}], \z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash e : \vp$\` by Lemma~\ref{lem:onedelta} (4)\\
    (6) $\G, \z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash e : \vp$\` by (3), resources in $\G$ with $\irr{\D}$ (parts) in\\\` the usage environment cannot be used in $e$\\
    (7) $\G, \z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash_{alt} \_ \to e :^z_{\ov{\D_i}} \s \Rightarrow \vp$\` by $Alt\_$\\
\end{tabbing}

@@

\item[Case:] $AltN_{\textrm{Not WHNF}}$
    We prove the theorem for constructing both an $AltN_{\textrm{Not WHNF}}$
        and an $AltN_{\textrm{WHNF}}$ from a proof-irrelevant $AltN_{\textrm{Not
        WHNF}}$, to prove the statement holds for any $\Rightarrow$ kind for
        $AltN$, rather than just $\Rrightarrow$ or $\Mapsto$.
\begin{tabbing}
    (1) $\G, \z[\irr{\D}]; \irr{\D}, \D' \vdash_{alt} K~\ov{\xo},\ov{y_i{:}_{1}\s_i}^n\to e :^z_{\irr{\D}} \s \Rrightarrow \vp$\\
    (2) $\G,\z[\irr{\D}],\ov{\xo},\ov{y_j{:}_{\D_j}\s_j}^n; \irr{\D}, \D' \vdash e : \vp$\`by inv. on $AltN_{\textrm{Not WHNF}}$\\
    (3) $\ov{\D_j} = \ov{\lctag{\irr{\D}}{K_j}}$\`by inv. on $AltN_{\textrm{Not WHNF}}$\\
    Subcase $\irr{\D}$ is consumed through $z$\\
    (4) $\G,\ov{\xo},\ov{y_j{:}_{\D_j[z/\irr{\D}]}\s_j}^n; \z[1], \D' \vdash e : \vp$\`by Lemma~\ref{lem:deltaone} (2,subcase)\\\`and vars in $\G$ with $\irr{\D}$ cannot be used by subcase\\
    Subcase constructing $AltN_{\textrm{WHNF}}$ ($\Mapsto$)\\
    (5) $\G,\ov{\xo},\ov{y_j{:}_{\D_i}\s_j}^n; \z[1], \D' \vdash e : \vp$\`by $\ov{y}$ is not used, and $Weaken$\\
    (6) $\G,\ov{\xo},\ov{y_j{:}_{\D_i}\s_j}^n, \z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash e : \vp$\`by Lemma~\ref{lem:onedelta} (5)\\\` and $\G$ vars do not mention $z$ by (4) \\
    (7) $\G, \z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash_{alt} K~\ov{\xo},\ov{y_i{:}_{1}\s_i}^n \to e :^z_{\ov{\D_i}} \s \Mapsto \vp$\`by $AltN_{\textrm{WHNF}}$ (6)\\
    Subcase constructing $AltN_{\textrm{Not WHNF}}$ ($\Rrightarrow$)\\
    (5) $\G,\ov{\xo},\ov{y_j{:}_{\lctag{\ov{\D_i}}{K_j}}\s_j}^n; \z[1], \D' \vdash e : \vp$\`by $\ov{y}$ does not consume resources\\
    (6) $\G,\ov{\xo},\ov{y_j{:}_{\lctag{\ov{\D_i}}{K_j}}\s_j}^n, \z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash e : \vp$\`by $1 \Rightarrow \D$ lemma (5)\\\` and $\G$ vars do not mention $z$ by (4) \\
    (7) $\G, \z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash_{alt} K~\ov{\xo},\ov{y_i{:}_{1}\s_i}^n \to e :^z_{\ov{\D_i}} \s \Rrightarrow \vp$\`by $AltN_{\textrm{Not WHNF}}$ (6)\\
    Subcase $\irr{\D}$ is (fully) consumed by $\ov{y}$ (after splitting)\\
    (4) $\G,\z[\ov{y}],\ov{\xo};\ov{y_j{:}_{1}\s_j}^n, \D' \vdash e : \vp$\`by Lemma~\ref{lem:deltaone} (2, subcase)\\\` and vars in $\G$ with $\irr{\D}$ cannot be used by subcase\\
    (5) $\G,\z[\ov{\D_i}],\ov{\xo};\ov{y_j{:}_{1}\s_j}^n, \D' \vdash e : \vp$\`by $Weaken$ and $z$ does not consume resources\\
    Subcase constructing $AltN_{\textrm{WHNF}}$ ($\Mapsto$)\\
    (6) $\G,\z[\ov{\D_i}],\ov{\xo},\ov{y_j{:}_{\D_i}\s_j}^n; \ov{\D_i}, \D' \vdash e : \vp$\`by Lemma~\ref{lem:onedelta} (5)\\\`and vars in $\G$ do not mention $\ov{y}$ (4,5)\\
    (7) $\G,\z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash K~\ov{\xo},\ov{y_i{:}_{1}\s_i}^n\to e :^z_{\ov{\D_i}} \s \Mapsto \vp$\`by $AltN_{\textrm{WHNF}}$\\
    Subcase constructing $AltN_{\textrm{Not WHNF}}$ ($\Rrightarrow$)\\
    (6) $\G,\z[\ov{\D_i}],\ov{\xo},\ov{y_j{:}_{\lctag{\ov{\D_i}}{K_j}}\s_j}^n; \ov{\D_i}, \D' \vdash e : \vp$\`by $1 \Rightarrow \D$ lemma (5)\\\`and vars in $\G$ do not mention $\ov{y}$ (4,5)\\
    (7) $\G,\z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash K~\ov{\xo},\ov{y_i{:}_{1}\s_i}^n\to e :^z_{\ov{\D_i}} \s \Rrightarrow \vp$\`by $AltN_{\textrm{Not WHNF}}$\\
\end{tabbing}
